---
title: "Clustering tracks: validation and real-world example"

vignette: >
  %\VignetteIndexEntry{Clustering tracks: validation and real-world example}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}

bibliography: references.bib

format:
  html:
    syntax-highlighting: github


execute:
  eval: false
  echo: true
  warning: false
  message: false
  collapse: true
  comment: "#>"
---


## NOTE
This vignette is work in progress!

## Validation with Pectoral Sandpiper ARGOS tracks

We validate `clusterTrack` using ARGOS Doppler telemetry tracks from male Pectoral Sandpipers (_Calidris melanotos_). These data are available in the `nomads` package as `pesa` and include curated use-site assignments from [@kempenaers2017breeding].  For more information, see `help(pesa)`. 
The dataset includes 103 individuals. Across tracks, there are 841 individual-level use sites in total. The median number of sites per individual is 6, with a range from 2 to 24 sites.

`clusterTrack` can be run on each individual track, and the inferred clusters can then be compared with an independent set of curated labels.

We use `clusterTrack.Vis` to save the interactive maps. Its `site()` function creates a thumbnail for each map and writes an `index.qmd` file, which can be edited and rendered into a local navigation page.

```{r}
pckgs <- c(
  "data.table",
  "nomads",
  "fs",
  "sf",
  "RSQLite",
  "clusterTrack",
  "clusterTrack.Vis",
  "foreach",
  "futurize",
  "aricode", # Efficient Computations of Standard Clustering Comparison Measures
  "clusterCrit" # clustering validation
)

sapply(pckgs, require, character.only = TRUE, quietly = TRUE)

plan(multisession)

interactive_map_dir <- "~/Desktop/pesa"
# or use paste0(tempdir(), "name")

if (dir_exists(interactive_map_dir)) {
  dir_delete(interactive_map_dir)
}
```

## Run `clusterTrack` and save interactive maps

The code below runs the full per-individual clustering workflow and collects the resulting `clusterTrack` output in `out`. These results are then used for validation by comparing the number of inferred clusters with the number of curated use sites, and by quantifying point-level agreement between inferred and curated labels.


```{r}
ids <- nomads::pesa[, .N, id]

o <- foreach(i = ids$id) %do%
  {
    x <- as_ctdf(nomads::pesa[id == i]) |>
      cluster_track()
    x[, id := i]
    m <- clusterTrack.Vis::map(
      x,
      fix_dateline = TRUE
    )

    m |> save_map(path = interactive_map_dir, name = i)

    x
  } |>
  futurize()

site(interactive_map_dir)

out <- rbindlist(o)
out[cluster == 0, cluster := NA]

smry <- foreach(i = seq_along(o)) %do%
  {
    x <- summary(o[[i]])
    set(x, j = "id", value = o[[i]]$id[1])
    x
  } |>
  rbindlist()
```

## Compare the number of known and inferred clusters

For each individual, the number of known clusters is compared with the number of clusters inferred by `clusterTrack`. 

```{r}
bench <- out[!is.na(site_id), .N, .(id, site_id)][, .N, id]
cltrk <- smry[, .N, id]
x <- merge(bench, cltrk, by = "id", suffixes = c("_old", "_new"))
x[, cor(N_old, N_new)]
```

## Modified Adjusted Rand index

The adjusted Rand index quantifies point-level agreement between the curated labels and the labels inferred by `clusterTrack`, after correcting for agreement expected by chance [@sundqvist2023mari] . 



```{r}
x <- out[!is.na(cluster) & !is.na(site_id)]

mari <- x[, .(v = MARI(cluster, site_id)), by = id]

mari[
  ,
  .(
    q05 = quantile(v, 0.05),
    median = median(v),
    mean = mean(v),
    q95 = quantile(v, 0.95)
  )
]
```

## Recall and precision

Recall and precision provide complementary summaries of point-level agreement between the curated labels and the inferred clusters.


```{r}
x <- out[!is.na(cluster) & !is.na(site_id)]

recall <- x[,
  .(v = extCriteria(cluster, site_id, "Recall")[[1]]),
  by = id
]

recall[
  ,
  .(
    q05 = quantile(v, 0.05),
    median = median(v),
    mean = mean(v),
    q95 = quantile(v, 0.95)
  )
]

precision <- x[, .(v = extCriteria(cluster, site_id, "Precision")[[1]]), by = id]

precision[
  ,
  .(
    q05 = quantile(v, 0.05),
    median = median(v),
    mean = mean(v),
    q95 = quantile(v, 0.95)
  )
]
```




## Notes on reproducibility

This vignette is intended as a record of the validation workflow, not as an automatically executed example. To reproduce the analysis, copy the code into a regular R script, install the optional workflow packages listed above, set `download_dir` to a suitable local directory, and run the script interactively.
