---
title: "Clustering tracks: validation and real-world example"

vignette: >
  %\VignetteIndexEntry{Clustering tracks: validation and real-world example}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}

bibliography: references.bib

format:
  html:
    syntax-highlighting: github


execute:
  eval: false
  echo: true
  warning: false
  message: false
  collapse: true
  comment: "#>"
---


Here we use an ARGOS Doppler telemetry dataset for male Pectoral Sandpipers (_Calidris melanotos_) to illustrate how `clusterTrack` works on a larger set of individual tracks. This dataset provides a useful validation case because it contains curated use-site assignments from [@kempenaers2017breeding].

`clusterTrack` is run independently on each individual track, and the extracted clusters are compared with the curated labels.
The companion package `clusterTrack.Vis` is used to save one interactive map per individual track and also to create a local `site` for easy navigation throughout the maps. 

For validation, the number of curated use sites is compared with the number of clusters inferred by `clusterTrack` and a  point-level agreement is quantified using several indices. 

```{r}
pckgs <- c(
  "data.table",
  "nomads",
  "fs",
  "sf",
  "RSQLite",
  "clusterTrack",
  "clusterTrack.Vis",
  "foreach",
  "futurize",
  "progressify",
  "aricode", # Efficient Computations of Standard Clustering Comparison Measures
  "clusterCrit" # clustering validation
)

sapply(pckgs, require, character.only = TRUE, quietly = TRUE)

plan(multisession)

interactive_map_dir <- paste0(tempdir(), "Kempenaers_and_Valcu_2017_pesa")

if (dir_exists(interactive_map_dir)) {
  dir_delete(interactive_map_dir)
}
```

## Run `clusterTrack` and save interactive maps

The validation data are available in the `R` package `nomads`.
The next block runs `clusterTrack` independently for each individual. The resulting maps are saved with the companion `clusterTrack.Vis` package.
`site()` creates a thumbnails for each map and an `index.qmd` file which can then be edited and compiled to create a navigation page. 
The cluster truck output is then gathered in `out`.

```{r, clusterTrack}
data(pesa)

ids <- pesa[, .N, id]

o <- foreach(i = ids$id) %do%
  {
    x <- as_ctdf(pesa[id == i]) |>
      cluster_track()

    m <- clusterTrack.Vis::map(
      x,
      fix_dateline = TRUE
    )

    m |> save_map(path = interactive_map_dir, name = i)

    x
  } |>
  progressify() |>
  futurize()

site(interactive_map_dir)

out <- rbindlist(o)
out[cluster == 0, cluster := NA]
```

## Compare the number of known and inferred clusters

For each individual, the number of known clusters is compared with the number of clusters inferred by `clusterTrack`. 

```{r}
x <- merge(
  out[!is.na(site_id), .N, .(id, site_id)][, .N, id],
  out[!is.na(cluster), .N, .(id, cluster)][, .N, id],
  by = "id",
  suffixes = c("_old", "_new")
)

x[, cor(N_old, N_new)]
```

## Modified Adjusted Rand index

The adjusted Rand index quantifies point-level agreement between the curated labels and the labels inferred by `clusterTrack`, after correcting for agreement expected by chance [@sundqvist2023mari] . 



```{r}
x <- out[!is.na(cluster) & !is.na(site_id)]

mari <- x[, .(v = MARI(cluster, site_id)), by = id]

mari[
  ,
  .(
    q05 = quantile(v, 0.05),
    median = median(v),
    mean = mean(v),
    q95 = quantile(v, 0.95)
  )
]
```

## Recall and precision

Recall and precision provide complementary summaries of point-level agreement between the curated labels and the inferred clusters.


```{r}
x <- out[!is.na(cluster) & !is.na(site_id)]

z <- x[,
  .(v = extCriteria(cluster, site_id, "Recall")[[1]]),
  by = id
]
ok
recall <- z[
  ,
  .(
    q05 = quantile(v, 0.05),
    median = median(v),
    mean = mean(v),
    q95 = quantile(v, 0.95)
  )
]

z <- x[, .(v = extCriteria(cluster, site_id, "Precision")[[1]]), by = id]

precision <- z[
  ,
  .(
    q05 = quantile(v, 0.05),
    median = median(v),
    mean = mean(v),
    q95 = quantile(v, 0.95)
  )
]
```




## Notes on reproducibility

This vignette is intended as a record of the validation workflow, not as an automatically executed example. To reproduce the analysis, copy the code into a regular R script, install the optional workflow packages listed above, set `download_dir` to a suitable local directory, and run the script interactively.
